The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-06 kalman_walk_paper.tex
added paragraphs on quant gen and islands

\begin{abstract}
%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
%the underlying mechanism may have changed,
%since distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small mutational changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
% Treating an organism as a ``black box''
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
This characterization implies that selectively neutral directions in genotype space should be common
% there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of idependently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate inversely proportional the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

\plr{I removed the $0.1 N_e$ because that's only in the "metapopulation" case,
which is where you've probably already got subpopulations acting like species already.}

@@

Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
but viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.

@@

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid} looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.
\plr{missing any important references here?}

@@

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    Two pairs of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown for each in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown).
    Solid lines depict the distance of the $F_2$ to optimum.
}
\end{figure}

@@

Is there evidence that this is actually occurring?
System drift and network rewiring has been inferred across the tree of life \citep{dalal2017transcription, johnson2017rewiring},
and there is often significant regulatory variation segregating within populations.
Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}.
Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
%strongly reproductively incompatible species
genetically distinct species
may be nearly morphologically indistinguishable.

@@

\paragraph{Nonlinearity and model assumptions}
Of course, real regulatory networks are not linear dynamical systems.
Most notably, physiological limits put upper bounds on expression levels,
implying saturating response curves.
It remains to be seen how well these results carry over into real systems,
but the fact that most nonlinear systems can be locally approximated by a linear one
suggests our qualitative results may hold more generally.
Furthermore, nonidentifiability (which implies the existence of neutral directions)
is often found in practice in moderately complex models of biological systems. 
\plr{Josh, add citation to sensitivity analysis}

@@

\paragraph{Peter to add quant gen discussion} about non-normality.

%Maybe compare to \citet{fierst}?
%
%Do we talk about the $G$ matrix and mutational correlation somewhere?
%
%Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
%
%What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
%
%This has implications for genetic load also!!
%
%hominids; neanderthal load.

>>>>

\begin{abstract}
%It is known that even if a species' phenotype has remained unchanged over evolutionary time,
%the underlying mechanism may have changed,
%since distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype remains unchanged over evolutionary time, the underlying mechanism may have changed, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small mutational changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
% Treating an organism as a ``black box''
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
This characterization implies that selectively neutral directions in genotype space should be common
% there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of idependently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

@@

Nonidentifiability is a major barrier to mechanistic understanding of real systems, 
but viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
It has also been shown theoretically that substantial genetic variation for a polygenic trait can be maintained
by migration between geographical regions having different genetic bases for the same, optimal trait value \citep{phillips1996maintenance}.
\plr{the citation is "maintenance of polygenic variation via a migration-selection balance..."}

@@

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatari2015simple}) looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    A pair of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown; 
    some would be a similar distance to the other side of the optimal set).
    The distribution of $F_2$s is appropriate for mixed homozygotes
    if both traits have a simple, one-locus genetic basis,
    but there is variation within each population at that locus.
}
\end{figure}

@@

Is there evidence that this is actually occurring?
System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring},
and there is often significant regulatory variation segregating within populations.
Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}, and hybrid incompatibilities have been attributed to cryptic molecular divergence underlying conserved body plans \citep{gavin2013embryonic}. 
Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
%strongly reproductively incompatible species
genetically distinct species
may be nearly morphologically indistinguishable.

@@

\paragraph{Population isolates and genetic load}
Above we have mostly discussed the case of two separated populations of equal size.
In contrast, consider a small subset of a large population that is suddenly isolated.
It will have a large amount of genetic variance, thanks to its history,
but a small population size -- and so can drift quite quickly until it exhausts its genetic variation.
By this point, it could have accumulated substantial incompatibility with the parental population.
Similarly, in a species with substantial isolation by distance, geographically local systems drift
might lead to genetic load due only to recombination, much as in \citep{phillips1996maintenance}.

\paragraph{Nonlinearity and model assumptions}
Of course, real regulatory networks are not linear dynamical systems.
Most notably, physiological limits put upper bounds on expression levels,
implying saturating response curves.
It remains to be seen how well these results carry over into real systems,
but the fact that most nonlinear systems can be locally approximated by a linear one
suggests our qualitative results may hold more generally.
Furthermore, nonidentifiability (which implies the existence of neutral directions)
is often found in practice in moderately complex models of biological systems \citep{gutenkunst2007universally, piazza2008diverse}.

This simple quantitative genetics model we use above
has been shown to produce good predictions in many situations,
even when the substantial number of simplifying assumptions are violated \citep{burger1994distribution,turelli1994genetic}.
\plr{the citations are to "on the distribution of the mean and variance" by Burger and Lande; and the "what, me normal" paper of turelli and barton}
The calculations above should be fairly robust even to substantial deviations from normality.
A larger effect on these predictions seems likely due to 
correlations due to molecular constraint, genetic linkage, population structure, historical contingency and so forth.
Although such considerations would not change the qualitative predictions of this model,
their combined effects could substantially change the predicted rate of accumulation of incompatibilities.

@@

%Maybe compare to \citet{fierst}?
%
%hominids; neanderthal load.
